---
title: "Government Employment Figures by Region"
output:
  pdf_document: default
  html_notebook: default
---
The data set is available from the website [LMIP Gov AU](http://lmip.gov.au/default.aspx?LMIP/Downloads/ABSLabourForceRegion). 

It contains regional figures for population counts of employment per industry. 

```{r}
data <- read.csv("data/employment/SA4_regions_feb2017.csv", header=TRUE)

# need to fix the numbers
for(i in 4:9) {
  data[,i] <- as.numeric(gsub(",", "", data[,i]))
}
head(data)
str(data)

```
Initially we will investigate the industry by region, the field of interest is the employment by industry total. The data is currently in long format we need to convert it to wide format.

```{r}
data1 <- data[,c(1:4)]

colnames(data1) <- c("Region", "State", "Industry", "TotalCount")
data1 <- data1[data1$Region != "Australia",]

head(data1)
temp <- reshape(data1, idvar=c("Region", "State"), timevar=c("Industry"), direction="wide")


temp$Location <- paste(temp$State, temp$Region, sep=" ")
head(temp)
names(temp)
colnames(temp) <- c("Region", 
                 "State",
                 "AGRIC_FRST_FISH", 
                 "MINING",
                 "MANUF",
                 "UTILITIES",
                 "CONSTR",
                 "WSALE_TRADE",
                 "RETAIL_TRADE",
                 "ACC_FOOD_SRV",
                 "TRNS_POST_WHOUSE",
                 "INFO_MEDIA_TELEC",
                 "FIN_INS_SRV",
                 "RENT_HIRE_RE_SRV",
                 "PROF_SCI_TECH_SRV",
                 "ADM_SUP_SRV",
                 "PADMIN_SAFETY",
                 "EDU_TRAIN",
                 "HEALTH_SOC_ASSIST",
                 "ARTS_REC_SRV",
                 "OTHER_SRV",
                 "Location")
df1 <- data.frame(Location=temp$Location,
                  temp[,4:ncol(temp)-1])
head(df1)
```
Looking initially at the data we can check if it is multivariate normal, 

```{r}
require(MVN)
X <- df1[,2:ncol(df1)]
X <- scale(X)
mardiaTest(X, qqplot=TRUE)
hzTest(X)
roystonTest(X)
```

The test results suggest that the data is not multivariate normal, and this is also reinforced by the qqplot. However we can still perform ordination and some analysis.

Initially inspecting the data through principle components.

```{r}
row.names(X) <- df1$Location
df1.prcomp <- princomp(X, cor=TRUE)
df1.prcomp
df1.prcomp$loadings
df1.prcomp$var <- df1.prcomp$sdev^2
```

The amount of variance explained per component

```{r}
total <- sum(df1.prcomp$var)
df1.prcomp$percentVar <- df1.prcomp$var / total
plot(df1.prcomp$percentVar, type="b")
data.frame(component=1:length(df1.prcomp$var), variance=df1.prcomp$var, percent=round(df1.prcomp$percentVar, 4))
```

We note that the first component explains 99% of the variance, however in the biplot it is difficult to decipher.

Printing the biplot for industries, there are a large number of regions hence we can move the viewport in order to get a better view of the ordination.

```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(1,2))
```

While the 1st component provides the most variation explained, it is difficult to visualise, the 2nd and 3rd components provide a separate axes which permits some simplication in the visualisation.

```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(2,3), expand=1)
```

Each segment of the axes is drawn separately in the series below.

```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(2,3), expand=1.4, xlim=c(-0.7, 0.0), ylim=c(0.0, 0.5))
```




```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(2,3), expand=3, xlim=c(0.0, 0.5), ylim=c(0.0, 0.4))
```

```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(2,3), expand=1.3, xlim=c(-0.1, 0.1), ylim=c(-0.1, 0.1))
```

```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(2,3), expand=1.5, xlim=c(-0.3, 0.0), ylim=c(-0.21, 0.0))
```




```{r, fig.width=10, fig.height=15}
biplot(df1.prcomp, choices=c(2,3), expand=1.4, xlim=c(-0.05, 0.21), ylim=c(-0.21, 0.0))
```


