---
title: "PCA Analysis of ABS Labour Force per Region"
output: html_notebook
---

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
require(ggplot2)
require(MVN)
require(scatterD3)
require(rbokeh)
```

The data set is available from the website [LMIP Gov AU](http://lmip.gov.au/default.aspx?LMIP/Downloads/ABSLabourForceRegion). 

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

data <- read.csv("data/employment/SA4_regions_feb2017.csv", header=TRUE)

# need to fix the numbers
for(i in 4:9) {
  data[,i] <- as.numeric(gsub(",", "", data[,i]))
}

data1 <- data[,c(1:4)]

colnames(data1) <- c("Region", "State", "Industry", "TotalCount")
data1 <- data1[data1$Region != "Australia",]

head(data1)
temp <- reshape(data1, idvar=c("Region", "State"), timevar=c("Industry"), direction="wide")


temp$Location <- paste(temp$State, temp$Region, sep=" ")
names(temp)
colnames(temp) <- c("Region", 
                 "State",
                 "AGRIC_FRST_FISH", 
                 "MINING",
                 "MANUF",
                 "UTILITIES",
                 "CONSTR",
                 "WSALE_TRADE",
                 "RETAIL_TRADE",
                 "ACC_FOOD_SRV",
                 "TRNS_POST_WHOUSE",
                 "INFO_MEDIA_TELEC",
                 "FIN_INS_SRV",
                 "RENT_HIRE_RE_SRV",
                 "PROF_SCI_TECH_SRV",
                 "ADM_SUP_SRV",
                 "PADMIN_SAFETY",
                 "EDU_TRAIN",
                 "HEALTH_SOC_ASSIST",
                 "ARTS_REC_SRV",
                 "OTHER_SRV",
                 "Location")
df1 <- data.frame(Location=temp$Location,
                  temp[,4:ncol(temp)-1])

X <- df1[,2:ncol(df1)]
X <- scale(X)

row.names(X) <- df1$Location
df1.prcomp <- princomp(X, cor=TRUE)
df1.prcomp$var <- df1.prcomp$sdev^2
```

## Principle components by variance.

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
total <- sum(df1.prcomp$var)
df1.prcomp$percentVar <- df1.prcomp$var / total

components <- data.frame(component=1:length(df1.prcomp$var), variance=df1.prcomp$var, percent=round(df1.prcomp$percentVar, 4), cumulativePercent=round(cumsum(df1.prcomp$percentVar),4))

components
```

The first five components explain > 95% of the variation, with the first nine components explaining 99% of the variance.

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
ggplot(components) +
  geom_bar(aes(x=component,y=percent),stat="identity")
```

The loadings for each of the industries are shown below.

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
df1.prcomp$loadings
```
The following plot provides an ordination of the loadings for locations against the industry weightings as the directions of
the principle components.

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

# plot the first two dimensions.
choices <- c(2,4)

pcobj <- df1.prcomp
nobs.factor <- sqrt(pcobj$n.obs)
d <- pcobj$sdev
u <- sweep(pcobj$scores, 2, 1 / (d * nobs.factor), FUN = '*')
v <- pcobj$loadings
df.u <- as.data.frame(sweep(u[,choices], 2, d[choices], FUN='*'))
 # Directions
  v <- sweep(v, 2, d, FUN='*')
  df.v <- as.data.frame(v[, choices])

  names(df.u) <- c('xvar', 'yvar')
  names(df.v) <- names(df.u)
 
C <- data.frame(x=df.v[,1],
                y=df.v[,2])
O <- data.frame(x=rep(0,nrow(C)),
                y=rep(0,nrow(C)))


figure(width=900,height=900) %>%
  ly_points(x=df.u[,1], y=df.u[,2],
            alpha=0.5) %>%
  ly_text(x=df.u[,1], y=df.u[,2],
          text=rownames(df.u),
          font_size="10pt") %>%
  ly_text(x=df.v[,1], y=df.v[,2],
          text=rownames(df.v), col="red") %>%
  ly_segments(x0=O$x, y0=O$y,
              x1=df.v[,1],
              y1=df.v[,2],
              col="red")
```
